\chapter{Questionnaire}
	\label{app.questionnaire}
	
	\section*{Personal information}
		\begin{multicols}{2}
		\begin{itemize}
			\item How old are you?
			\item What is your level of education? 
			\item What is your first language?
			\item Did you grow up speaking any other languages?
			\item What do you do for a living?
			\item Were you born in Liverpool?
			\item If yes, in which district?
			\item If not, where?
			\item How old were you when you came to Liverpool?
			\item In which district do you live now?
			\item In how many other places besides Liverpool have you lived?
			\item Was your mother born in Liverpool?
			\item If not, where?
			\item What's her profession?
			\item Was your father born in Liverpool?
			\item If not, where?
			\item What's his profession?
			\item Would you consider yourself WC or MC?
			\item What is your creed?
			\item If Christian, protestant or catholic?
		\end{itemize}
	\end{multicols}
	
	\section*{Local knowledge}
	\begin{multicols}{2}
		\begin{itemize}
			\item Do you know who is mayor of Liverpool (Joe Anderson)?
			\item Do you know who controls the City Council (Labour)?
			\item Do you vote?
			\item Please name as many quarters/districts of Liverpool as you can.
		\end{itemize}
	\end{multicols}
	
	\section*{Sports and hobbies}
	\begin{multicols}{2}
		\begin{itemize}
			\item Are you a football fan? Liverpool FC or Everton FC?
			\item Does this matter a lot? Would you say the two clubs are intrinsically different in character? Does that also hold for their supporters?
			\item What sports do people prefer here (apart from football)? Do you play football/other sport?
		\end{itemize}
	\end{multicols}
	
	\section*{``Children’s lore''}
	\begin{multicols}{2}
		\begin{itemize}
			\item Could you describe the district you grew up in? How do I have to picture the environments?
			\item If applicable: Do you still live there? If not, how about the new district?
			\item Can you tell me about some games you played, in the school yard for example?
			\item Do you know ``conkers''? How does that work?
			\item Did you use any rhymes to make fun of people?
			\item Do you remember any counting out rhymes you used?
			\item Did you know any clapping games? What songs went with them?
			\item Did you ever get into fights when you were a kid?
			\item Did you have any rules about what was considered fair?
			\item Was your class ethnically mixed? How did you experience that? Any problems?
			\item Was it a ``close-knit'' community in your district where everyone knew everyone? A good place to grow up?
			\item Do you remember what you wanted to do for a living when you were young? Where did you get that idea and why was it attractive?
		\end{itemize}
	\end{multicols}
	
	\section*{Attachment to Liverpool}
	\begin{multicols}{2}
		\begin{itemize}
			\item Do you like living here?
			\item Are there places you would rather live?
			\item Do you think Liverpool is a good place to raise kids? A place where they will have a good life?
			\item Would you say most of your best friends are from Liverpool or from outside the city?
			\item What are the first things that come to mind when you think of Liverpool?
			\item What do you like about Liverpool?
			\item What don't you like about Liverpool?
			\item If someone came to visit the city what would you show them?
			\item Where can you experience the \enquote{real} Liverpool?
			\item Are there any festivals or traditions in Liverpool you can tell me about?
			\item Is there a Liverpudlian sense of humour? What is it like?
		\end{itemize}
	\end{multicols}
	
	\section*{Identity}
	\begin{multicols}{2}
		\begin{itemize}
			\item What would you call yourself (British, English, Northerner, Merseysider, Liverpudlian, Scouser)? If several, in what order?
			\item How would you feel if someone called you a Scouser?
			\item What are Woolybacks? Where do they live?
			\item Plastic Scousers?
			\item Scally Scouser?
			\item ``In the North, but not of it'' --- your opinion?
		\end{itemize}
	\end{multicols}
	
	\section*{Recent developments}
	\begin{multicols}{2}
		\begin{itemize}
			\item Do you think Liverpool has changed in the last 20 years? For better or worse?
			\item What do you think of Liverpool One?
			\item Do you think urban restructuring is destroying Liverpool's character?
			\item Do you have the impression that Liverpool's economy is recovering?
			\item Suppose you were Liverpool's mayor. What would be your first project?
			\item What did you think of Liverpool as European Capital of Culture?
			\item Do you, as a Liverpudlian, feel neglected by the central government? If yes, in what ways?
			\item More and more tourists seem to be visiting Liverpool. Is that a good thing?
			\item How do you feel about the Beatles cult?
		\end{itemize}
	\end{multicols}
	
	\section*{Liverpool's image}
	\begin{multicols}{2}
		\begin{itemize}
			\item Are you familiar with the label ``Livercool''? Do you think it is appropriate?
			\item Do you think Liverpool is stigmatised in England/the United Kingdom?
			\item Why do you think some people have rather negative attitudes towards Liverpool?
			\item Older subjects: Do you remember the Toxteth riots of 1981? How did you experience them? What exactly happened?
			\item Do you think the riots still form part of Liverpool's image in the rest of the country?
			\item ``The world in one city'' - would you say this is still true today?
			\item Would you say Liverpool is a dangerous place? More dangerous than others? What are dangerous areas?
			\item Liverpool frequently ``doubles'' for other cities in films. Does that make you proud or rather angry?
		\end{itemize}
	\end{multicols}
	
	\section*{Liverpool and Manchester}
	\begin{multicols}{2}
		\begin{itemize}
			\item There's an old rivalry between Liverpool and Manchester. Is that still alive today? Why do you think that is?
			\item Do you think Manchester is doing better than Liverpool? Why (not)?
			\item Do you sometimes go to Manchester? What for?
			\item What are the people there like? Can you characterise them?
		\end{itemize}
	\end{multicols}
	
	\section*{Reading passage}

	\section*{Word list}

	\section*{Imitation}

	\section*{Questions about Scouse}
	\begin{multicols}{2}
		\begin{itemize}
			\item Can you tell someone is from Liverpool by the way they talk? How?
			\item What would you say are the most typical features of Liverpool English? Certain words or sounds?
			\item What do you think of the sound of Scouse? Pleasant, likeable, friendly, uneducated, rough...?
			\item Do you think Scouse has changed in the last 20 years? How? 
			\item If yes, how do you evaluate this change?
		\end{itemize}		
	\end{multicols}